---
title: "Week 1 Assignment – Loading Data into a Data Frame"
author: "Naomi Buell"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# load packages
library(ggplot2)
library(dplyr)
library(tidyverse)
library(janitor)
library(patchwork)
library(forcats)
```

# Presidential Primary Election Polls
## Introduction
For this assignment, I chose the [current presidential primary election poll data from FiveThirtyEight]( https://projects.fivethirtyeight.com/polls/president-primary-r/2024/national/). Here is an 538/ABC news article associated with this data called ["This could be the shortest presidential primary ever"](https://abcnews.go.com/538/shortest-presidential-primary/story?id=106819031). The article discusses the 2024 Republican presidential nomination race, and explains that it will likely be ending quickly due to Trump's strong support within the party compared to Nikki Haley. In particular, Nikki Haley is losing in her home state of South Carolina, where Trump leads by over 30 percentage points.

## Data transformation
Here, I load the data from the source online and revise some variable data types.
```{r load-data, eval = TRUE}
# Loading data (original data file is accessible through code) and transforming data types
primaries <- read_csv("https://projects.fivethirtyeight.com/polls/data/president_primary_polls.csv", 
               col_types = cols(start_date = col_date("%m/%d/%y"), end_date = col_date("%m/%d/%y")), # changing char to date data types when loading
                 na = c(""))
```

I rename variables to be more straightforward, further revise variable types, and subset the data to just this election cycle's Republican primaries. 
```{r rename-subset-data, eval = TRUE}
# rename and subset data
primaries_cleaned <- primaries |> 
  janitor::clean_names() |>
  select(fte_grade, methodology, transparency_score, state, start_date, end_date, sample_size, party, answer, pct) |>
  filter(start_date >= ymd(20230101) & party == "REP") |>  
  rename(five_thirty_eight_grade = fte_grade, percent_vote = pct, candidate = answer) |> 
  mutate(state = fct(state),
         party = fct(party),
         five_thirty_eight_grade = fct(five_thirty_eight_grade),
         methodology = fct(methodology),
         candidate = fct(candidate))
```

Here is a peak at the cleaned dataset:
```{r view-data, eval = TRUE}
# browse data
head(primaries_cleaned)
```

## Data Visualization
Taking a closer look at just the final 3 republican candidates over time:
```{r plot-data, eval = TRUE}
# Plotting just the last 3 republican presidential primary candidates to remain in the primaries (Donald Trump, Nikki Haley, and Ron DeSantis:
primaries_filtered <- dplyr::filter(primaries_cleaned, candidate %in% c("Trump", "Haley", "DeSantis"))
primaries_filtered$sc <- if_else(primaries_filtered$state == "South Carolina", "South Carolina", "National", "National") # Adding dummy var for SC, assuming NAs are not SC, and using strings so they show up plot.

ggplot(
  data = primaries_filtered,
  mapping = aes(x = end_date, y = percent_vote, color = candidate)
) +
  geom_point(size = 0.2) +
  geom_smooth() +
  labs(
    x = "Date",
    y = "Percent of the poll",
    title = "Republican Primaries on the whole and in South Carolina",
    color = "Candidate"
 ) +
facet_wrap(~ sc)
```

## Conclusion
After loading the primary data and transforming it to reflect the current republican primaries, I verified that Trump is winning the primaries with Haley in second--in South Carolina (SC) and across the country. Although Trump's polls are lower in SC than in the rest of the country, it does look like he is leading in the South Carolina primaries by 30 percentage points like the article states.
